### 2022-05-09 Neutrino.tex
rewrite with new class

\newcommand\MD[1][]{M_{\mathrm{D}#1}}
\newcommand\MN[1][]{M_{N#1}}
\newcommand{\nRHN}{n_{\text{RHN}}}

@@

\subsection{Models}
\paragraph{Dirac neutrino model}
The simplest model for neutrino masses are given by extending the SM Yukawa \eqref{eq:SMyukawa} with ($\nRHN$ copies of) right-handed neutrino $N$:
\begin{align}
   \mathcal L\w{Dirac-$\nu$}&=
\left(  \overline{U}\Yu H \PL Q
- \overline{D}\Yd H^\dagger \PL Q
+ \overline{N}\Yn H \PL L
- \overline{E}\Ye H^\dagger \PL L
\right)+ \text{h.c.}
\\&=
\left(- \overline{Q^a}\Yu^\dagger \epsilon^{ab}H^{b*} \PR U
- \overline{Q^a}\Yd^\dagger H^{a} \PR D
- \overline{L^a}\Yn^\dagger \epsilon^{ab}H^{b*} \PR N
- \overline{L^a}\Ye^\dagger H^{a} \PR E
\right) + \text{h.c.}
\end{align}
Neutrinos become Dirac fermions and the discussion goes parallel to the CKM matrix:
\begin{equation}
 \Yn = U_n Y_n\wdiag V_n^\dagger,\qquad
\{\nuL, \eL, \overline N, \overline E\}^{\text{mass basis}} = \{V_n^\dagger L^1, V_e^\dagger L^2, \overline N U_n, \overline E U_e\},
\end{equation}
where  $\Yn\in\mathbb C^{\nRHN\times3}$, $V_n\in\mathbb U_{\mathbb C}^3$, and $U_n\in\mathbb U_{\mathbb C}^{\nRHN}$.
The Lagrangian is now given by
\begin{align}
 \mathcal L&\supset\overline{L}\ii\gamma^\mu(-\ii g_2 W_\mu)\PL L
\supset
\frac{g_2}{\sqrt2}\left[
\overline{L^1} \slashed W^+\PL L^2
+\overline{L^2}\slashed W^-\PL L^1
\right]
\\&=
\frac{g_2}{\sqrt2}\left[
\overline{(L^1)^{\text{mass}}} V_n^\dagger V_e\slashed W^+\PL (L^2)^{\text{mass}}
+\overline{(L^2)^{\text{mass}}}V_e^\dagger V_n\slashed W^-\PL(L^1)^{\text{mass}}
\right]
\\&=
\frac{g_2}{\sqrt2}\left[
\overline{\nuL^{\text{mass}}} U^\dagger\slashed W^+\eL^{\text{mass}}
+\overline{\eL^{\text{mass}}}U\slashed W^-\nuL^{\text{mass}}
\right],
\end{align}
where the definition \eqref{eq:U-defining} provides the equality
\begin{equation}
 U=V_e^\dagger V_n\equiv U\w{PMNS}.
\end{equation}
According to our terminology, this $3\times 3$ unitary matrix $U$ is called ``the PMNS matrix'' for this model.
Note that this is given in \emph{the opposite manner} to the CKM matrix ($V\w{CKM}=V_u^\dagger V_d$).

The mass eigenstate Dirac fields are given by
\begin{equation}
 \nu^{\text{mass}} = \pmat{\nuL^{\text{mass}}\\N^{\text{mass}}} = 
\nuL^{\text{mass}}+N^{\text{mass}}=
V_n^\dagger L^1 + U_n^\dagger N
\qquad
 \Bigl(L^1 = \PL V_n \nu^{\text{mass}},~
 N = \PR U_n \nu^{\text{mass}}\Bigr).
\end{equation}
Note that $\nRHN=\{2,3\}$ because models with $\nRHN<3$ yields $3-\nRHN$ massless left-handed neutrinos, while $\nRHN>3$ results in $\nRHN-3$ massless right-handed neutrinos.

\paragraph{Type-I see-saw}
The type-I see-saw models are given by 
\begin{align}
   \mathcal L_{\text{type-I}}&=
\left(  \overline{U}\Yu H \PL Q
- \overline{D}\Yd H^\dagger \PL Q
+ \overline{N}\Yn H \PL L
- \overline{E}\Ye H^\dagger \PL L
- \frac{1}{2}\overline{N}\MN N^\cc
\right)+ \text{h.c.}
\\&=
\left(- \overline{Q^a}\Yu^\dagger \epsilon^{ab}H^{b*} \PR U
- \overline{Q^a}\Yd^\dagger H^{a} \PR D
- \overline{L^a}\Yn^\dagger \epsilon^{ab}H^{b*} \PR N
- \overline{L^a}\Ye^\dagger H^{a} \PR E
- \frac{1}{2}\overline{N}\MN N^\cc\right) + \text{h.c.},
\label{eq:type-1-lagrangian}
\end{align}
where $M_N$ is a $\nRHN\times\nRHN$ complex symmetric matrix.
Models with $\nRHN\ge2$ are viable, because $\nRHN<3$ gives $3-\nRHN$ massless (Weyl) neutrinos and $2\nRHN$ massive (Majorana) neutrinos, while all neutrinos acquire Majorana mass  if $\nRHN\ge 3$.

We introduce left-handed Weyl spinors $\xi$ and $\chi$ to avoid notational confusion and diagonalize the Dirac mass term $\MD$:
\begin{equation}
 L^1 \equiv \pmat{\xi\\0}\equiv \pmat{\nuL\\0},\quad
 N   \equiv \pmat{0\\\bar\chi}\equiv \pmat{0\\\nR},\quad
 \MD := \frac{v}{\sqrt2}Y_n,\quad
 \MD\wdiag := U_n \MD Y_n^\dagger.
\end{equation}
Then the mass term is written in matrix form:
\begin{align}
\label{eq:MajoranaNuLag}
 \mathcal L_{\text{type-I}}
\supset
 -\frac{v}{\sqrt2}\overline{N}Y_n \PL L^1 - \frac{1}{2}\overline{N}\MN N^\cc + \text{h.c.}
=-\frac12\pmat{\xi & \chi}\pmat{0 & \MD^\TT \\ \MD & \MN} \pmat{\xi \\ \chi} + \text{h.c.} =: -\frac12\tilde\nu^\TT \tilde M\tilde \nu+\text{h.c.},
\end{align}
where tildes denote ``larger'' objects. We can AT-diagonalize $\tilde M$, which is a symmetric $\mathbb{C}^{(3+\nRHN)\times(3+\nRHN)}$ matrix, as usual:
\begin{equation}
  \tilde M = \tilde R \tilde M\wdiag \tilde R^\TT;
\qquad
-\mathcal L\w{Dirac-$\nu$}\supset
\frac12\tilde \nu^\TT \tilde M\tilde\nu
=\frac12(\tilde\nu^{\text{mass}})^\TT \tilde M\wdiag \tilde\nu^{\text{mass}};\qquad \tilde\nu^{\text{mass}} =\tilde R^\TT\tilde\nu.
\end{equation}
Mass eigenstates $\nu^{\text{mass}}$ are Majorana fermions given by left-handed Weyl spinors.
The neutrino mixing are given by
\begin{equation}
 \pmat{\xi\\\chi} \equiv \pmat{\nuL\\(\nR)^\dagger} = \tilde R^*\pmat{\nu_{1\TO3} \\ \nu_{4\TO}},\qquad
 \therefore
   U = \pmat{V_e^\dagger & 0 \\ 0 & X}\tilde R^*,
\label{eq:type-I-mixing}
\end{equation}
where $Y_e$ comes from the charged lepton mass diagonalization, while $X$ is a unitary matrix (but unphysical if, as usual, the right-handed neutrinos are indistinguishable).

\paragraph{Connection of the above two models}
If we set $\MN=0$ in the Type-I see-saw ($\nRHN=3$), the AT-diagonalization becomes
\begin{align}
\tilde M = \frac12\pmat{0 & (U_n\MD\wdiag V_n^\dagger)^\TT \\ U_n\MD\wdiag V_n^\dagger & 0},\quad
  \tilde R=\pmat{V_n^* & \ii V_n^* \\ U_n & -\ii U_n},\quad
  \tilde M\wdiag=\pmat{\MD\wdiag & 0 \\ 0 & \MD\wdiag}.
\end{align}
As three pairs of degenerate Weyl fermions form three Dirac neutrinos, the Dirac neutrino model is reproduced as expected.

@@

The matrix $U'$ is theoretically given with some matrix $\Theta$ by~\cite{Blennow:2011vn}\footnote{%
\TODO{}proof? or obvious if considers $\gSU(6)/(\gSU(3)\times\gSU(3))$?}
\begin{equation}
 U'\equiv \exp\pmat{0 & -\Theta^\dagger \\ \Theta & 0}
=\sum_{n=0}^\infty\pmat{
  {(-\Theta^\dagger\Theta)^{n}}/{(2n)!} &
  {-(-\Theta^\dagger\Theta)^{n}\Theta^\dagger}/{(2n+1)!}\\
  {+\Theta(-\Theta^\dagger\Theta)^{n}}/{(2n+1)!}&
  {(-\Theta\Theta^\dagger)^{n}}/{(2n)!}
}
\end{equation}
but practically we derive it in perturbative expansion.
For example, the type-I case is evaluated as\footnote{Cf.~\texttt{calculator/neutrino/diagonalization\_perturbative.wl}}
\begin{align}
 \text{(diag.)} = U^\TT\pmat{0 & \MD^\TT \\ \MD & \MN} U;\qquad
&U'\approx\pmat{
  1-\frac12 \Theta_0^\dagger \Theta_0&
  -\Theta_0^\dagger\\
 \Theta_0&
  1-\frac12\Theta_0\Theta_0^\dagger
};\quad \Theta_0:=-\MN^{-1}\MD,
\label{eq:typeIseesaw-blockdiag-1}\\
&\text{(block diag.)}=U'^\TT\pmat{0 & \MD^\TT \\ \MD & \MN} U'\approx\pmat{
  -\MD^\TT\MN^{-1}\MD & 0 \\ 0 & \MN
},
\label{eq:typeIseesaw-blockdiag-2}
\end{align}
where however note that $\Theta_0$ is close but not equal to $\Theta$.

@@

\subsection{PMNS matrix}\label{sec:PMNS}
As experiments have not found any deviation from $[U]\w{upper left}$ being unitary, we express the mixing with the PMNS matrix
 \GRAY{(Pontecorvo--\japanese{牧}--\japanese{中川}--\japanese{坂田})}
parameterized by
\begin{equation}
\begin{split}
   U\w{PMNS}&=
 \pmat{1&&\\&\co{23}&\si{23}\\&-\si{23}&\co{23}}
 \pmat{\co{13}&&\si{13}\ee^{-\ii\delta\w{CP}}\\&1&\\-\si{13}\ee^{\ii\delta\w{CP}}&&\co{13}}
 \pmat{\co{12}&\si{12}&\\-\si{12}&\co{12}&\\&&1}
\pmat{\ee^{\ii\eta_1} \\ & \ee^{\ii\eta_2}\\&&1}
 %
\\&=
 \pmat{
 \co{12} \co{13} & \si{12} \co{13} & \si{13} \ee^{-\ii\delta\w{CP}}\\
 -\si{12} \co{23} - \co{12} \si{23} \si{13} \ee^{\ii\delta\w{CP}}& \co{12} \co{23} - \si{12}\si{23}\si{13} \ee^{\ii\delta\w{CP}}& \si{23}\co{13}\\
  \si{12}\si{23} - \co{12} \co{23} \si{13}\ee^{\ii\delta\w{CP}} & -\co{12}\si{23}-\si{12}\co{23}\si{13}\ee^{\ii\delta\w{CP}} & \co{23} \co{13}
}\pmat{\ee^{\ii\eta_1} \\ & \ee^{\ii\eta_2}\\&&1},
\end{split}
\end{equation}
where $\theta_{ij}\in[0,\pi/2]$ and $\delta\w{CP}\in[0,2\pi]$ as shown in \cref{eq:GeneralUnitary33}.
In the Dirac neutrino model, $\eta_1$ and $\eta_2$ (``Majorana phases'') are unphysical because we can rotate $\nR$ to eliminate them, while the rotation is not allowed in presence of the Majorana mass term.

@@

It also agrees with NuFIT~\cite[v5.0]{NUFIT} (their convention is given in Ref.~\cite{Esteban:2018azc}) except for the Majorana phases,
$
 \eta_i = \CONV{\alpha_i}.
$

\subsection{Casas--Ibarra parameterization}
We start from the approximation given in Eq.~\eqref{eq:typeIseesaw-blockdiag-2}:
\begin{equation}
 m\w L \approx -U\w{PMNS}^\TT \MD^\TT \MN^{-1}\MD U\w{PMNS},\qquad
 m\w H \approx U_X^\TT \MN U_X,
\end{equation}
where $m\w L$ and $m\w H$ are diagonal mass matrices for three light neutrinos and heavier neutrinos.
We then obtain
\begin{equation}
  m\w L
\approx -U\w{PMNS}^\TT \MD^\TT (U_X^* m\w H U_X^\dagger)^{-1}\MD U\w{PMNS}
=       -U\w{PMNS}^\TT \MD^\TT U_Xm\w H^{-1}U_X^\TT\MD U\w{PMNS}
\end{equation}
and decompose it as
\begin{equation}
   [\ii m\w L]^{1/2}[\ii m\w L]^{1/2}
\approx     \left(\frac{v}{\sqrt2}m\w H^{-1/2}U_X^\TT Y_n U\w{PMNS}\right)^{\TT} \left(\frac{v}{\sqrt2} m\w H^{-1/2}U_X^\TT Y_n U\w{PMNS}\right),
\end{equation}
which is the master equation for Casas-Ibarra parameterization~\cite{Casas:2001sr}.
Note that $U_X$ is unphysical, i.e., we can fix the basis of $\nR$ so that $U_X=1$.
This basis is equivalent (under our approximation) to the basis in which $\MN$ becomes diagonal.

\paragraph{Example: three right-handed neutrinos}
Let us assume all the neutrinos are massive thanks to three right-handed neutrinos. Then $M\w L\wdiag$ is invertible and
\begin{equation}
 R:=-\ii m\w H^{-1/2}U_X^\TT \MD  U\w{PMNS}  m\w L^{-1/2}
\quad\Longrightarrow\quad R^\TT R\approx1.
\end{equation}
Conversely, with a matrix $R$ satisfying $R^\TT R=1$, the Yukawa matrix is given by
\begin{equation}
 \frac{v}{\sqrt2}Y_n
  \approx \ii U_X^*\sqrt{m\w H}R\sqrt{m\w L}U\w{PMNS}^\dagger
\end{equation}
or
\begin{equation}
 \frac{v}{\sqrt2}Y_n\Big|\w{$\MN$ being diagonal}
  \approx \ii \sqrt{m\w H}R\sqrt{m\w L}U\w{PMNS}^\dagger.
\end{equation}

@@

With this parameterization, the neutrino mixing matrix is given by, in the basis with $\MN$ being diagonal,
\begin{equation}
 U \approx \pmat{
  U\w{PMNS}
  &  \MD^\dagger \MN^{*-1} U_X
  \\ -\MN^{-1} \MD U\w{PMNS}
  &  U_X
}
\approx\pmat{U\w{PMNS}&0\\0&1}\pmat{
  1
  &  -(-\ii{m\w H^{-1/2}}R{m\w L^{1/2}})^\dagger
  \\ -\ii m\w H^{-1/2}R{m\w L^{1/2}}
  &  1
}.
\end{equation}

>>>>

@@

\subsection{Models}
\paragraph{Dirac neutrino model}
The simplest model for neutrino masses are given by extending the SM Yukawa \eqref{eq:SMyukawa} with ($n\w{RHN}$ copies of) right-handed neutrino $N$:
\begin{align}
   \mathcal L\w*{Dirac-$\nu$}&=
\left(  \overline{U}\Yu H \PL Q
- \overline{D}\Yd H^\dagger \PL Q
+ \overline{N}\Yn H \PL L
- \overline{E}\Ye H^\dagger \PL L
\right)+ \text{h.c.}
\\&=
\left(- \overline{Q^a}\Yu^\dagger \epsilon^{ab}H^{b*} \PR U
- \overline{Q^a}\Yd^\dagger H^{a} \PR D
- \overline{L^a}\Yn^\dagger \epsilon^{ab}H^{b*} \PR N
- \overline{L^a}\Ye^\dagger H^{a} \PR E
\right) + \text{h.c.}
\end{align}
Neutrinos become Dirac fermions and the discussion goes parallel to the CKM matrix:
\begin{equation}
 \Yn = U_n Y_n\wdiag V_n^\dagger,\qquad
\{\nuL, \eL, \overline N, \overline E\}^{\text{mass basis}} = \{V_n^\dagger L^1, V_e^\dagger L^2, \overline N U_n, \overline E U_e\},
\end{equation}
where  $\Yn\in\mathbb C^{n\w{RHN}\times3}$, $V_n\in\mathbb U_{\mathbb C}^3$, and $U_n\in\mathbb U_{\mathbb C}^{n\w{RHN}}$.
The Lagrangian is now given by
\begin{align}
 \mathcal L&\supset\overline{L}\ii\gamma^\mu(-\ii g_2 W_\mu)\PL L
\supset
\frac{g_2}{\sqrt2}\left[
\overline{L^1} \slashed W^+\PL L^2
+\overline{L^2}\slashed W^-\PL L^1
\right]
\\&=
\frac{g_2}{\sqrt2}\left[
\overline{(L^1)^{\text{mass}}} V_n^\dagger V_e\slashed W^+\PL (L^2)^{\text{mass}}
+\overline{(L^2)^{\text{mass}}}V_e^\dagger V_n\slashed W^-\PL(L^1)^{\text{mass}}
\right]
\\&=
\frac{g_2}{\sqrt2}\left[
\overline{\nuL^{\text{mass}}} U^\dagger\slashed W^+\eL^{\text{mass}}
+\overline{\eL^{\text{mass}}}U\slashed W^-\nuL^{\text{mass}}
\right],
\end{align}
where the definition \eqref{eq:U-defining} provides the equality
\begin{equation}
 U=V_e^\dagger V_n\equiv U\w{PMNS}.
\end{equation}
According to our terminology, this $3\times 3$ unitary matrix $U$ is called ``the PMNS matrix'' for this model.
Note that this is given in \emph{the opposite manner} to the CKM matrix ($V\w{CKM}=V_u^\dagger V_d$).

The mass eigenstate Dirac fields are given by
\begin{equation}
 \nu^{\text{mass}} = \pmat{\nuL^{\text{mass}}\\N^{\text{mass}}} = 
\nuL^{\text{mass}}+N^{\text{mass}}=
V_n^\dagger L^1 + U_n^\dagger N
\qquad
 \Bigl(L^1 = \PL V_n \nu^{\text{mass}},~
 N = \PR U_n \nu^{\text{mass}}\Bigr).
\end{equation}
Note that $n\w{RHN}=\{2,3\}$ because models with $n\w{RHN}<3$ yields $3-n\w{RHN}$ massless left-handed neutrinos, while $n\w{RHN}>3$ results in $n\w{RHN}-3$ massless right-handed neutrinos.

\paragraph{Type-I see-saw}
The type-I see-saw models are given by 
\begin{align}
   \mathcal L_{\text{type-I}}&=
\left(  \overline{U}\Yu H \PL Q
- \overline{D}\Yd H^\dagger \PL Q
+ \overline{N}\Yn H \PL L
- \overline{E}\Ye H^\dagger \PL L
- \frac{1}{2}\overline{N}M_N  N^\cc
\right)+ \text{h.c.}
\\&=
\left(- \overline{Q^a}\Yu^\dagger \epsilon^{ab}H^{b*} \PR U
- \overline{Q^a}\Yd^\dagger H^{a} \PR D
- \overline{L^a}\Yn^\dagger \epsilon^{ab}H^{b*} \PR N
- \overline{L^a}\Ye^\dagger H^{a} \PR E
- \frac{1}{2}\overline{N}M_N  N^\cc\right) + \text{h.c.},
\label{eq:type-1-lagrangian}
\end{align}
where $M_N$ is a $n\w{RHN}\times n\w{RHN}$ complex symmetric matrix.
Models with $n\w{RHN}\ge2$ are viable, because $n\w{RHN}<3$ gives $3-n\w{RHN}$ massless (Weyl) neutrinos and $2n\w{RHN}$ massive (Majorana) neutrinos, while all neutrinos acquire Majorana mass  if $n\w{RHN}\ge 3$.

We introduce left-handed Weyl spinors $\xi$ and $\chi$ to avoid notational confusion and diagonalize the Dirac mass term $M\w{D}$:
\begin{equation}
 L^1 \equiv \pmat{\xi\\0}\equiv \pmat{\nuL\\0},\quad
 N   \equiv \pmat{0\\\bar\chi}\equiv \pmat{0\\\nR},\quad
 M\w{D} := \frac{v}{\sqrt2}Y_n,\quad
 M\w{D}\wdiag := U_n M\w{D} Y_n^\dagger.
\end{equation}
Then the mass term is written in matrix form:
\begin{align}
\label{eq:MajoranaNuLag}
 \mathcal L_{\text{type-I}}
\supset
 -\frac{v}{\sqrt2}\overline{N}Y_n \PL L^1 - \frac{1}{2}\overline{N}M_N  N^\cc + \text{h.c.}
=-\frac12\pmat{\xi & \chi}\pmat{0 & M\w{D}^\TT \\ M\w{D} & M_N } \pmat{\xi \\ \chi} + \text{h.c.} =: -\frac12\tilde\nu^\TT \tilde M\tilde \nu+\text{h.c.},
\end{align}
where tildes denote ``larger'' objects. We can AT-diagonalize $\tilde M$, which is a symmetric $\mathbb{C}^{(3+n\w{RHN})\times(3+n\w{RHN})}$ matrix, as usual:
\begin{equation}
  \tilde M = \tilde R \tilde M\wdiag \tilde R^\TT;
\qquad
-\mathcal L\w*{Dirac-$\nu$}\supset
\frac12\tilde \nu^\TT \tilde M\tilde\nu
=\frac12(\tilde\nu^{\text{mass}})^\TT \tilde M\wdiag \tilde\nu^{\text{mass}};\qquad \tilde\nu^{\text{mass}} =\tilde R^\TT\tilde\nu.
\end{equation}
Mass eigenstates $\nu^{\text{mass}}$ are Majorana fermions given by left-handed Weyl spinors.
The neutrino mixing are given by
\begin{equation}
 \pmat{\xi\\\chi} \equiv \pmat{\nuL\\(\nR)^\dagger} = \tilde R^*\pmat{\nu_{1\TO3} \\ \nu_{4\TO}},\qquad
 \therefore
   U = \pmat{V_e^\dagger & 0 \\ 0 & X}\tilde R^*,
\label{eq:type-I-mixing}
\end{equation}
where $Y_e$ comes from the charged lepton mass diagonalization, while $X$ is a unitary matrix (but unphysical if, as usual, the right-handed neutrinos are indistinguishable).

\paragraph{Connection of the above two models}
If we set $M_N =0$ in the Type-I see-saw ($n\w{RHN}=3$), the AT-diagonalization becomes
\begin{align}
\tilde M = \frac12\pmat{0 & (U_nM\w{D}\wdiag V_n^\dagger)^\TT \\ U_nM\w{D}\wdiag V_n^\dagger & 0},\quad
  \tilde R=\pmat{V_n^* & \ii V_n^* \\ U_n & -\ii U_n},\quad
  \tilde M\wdiag=\pmat{M\w{D}\wdiag & 0 \\ 0 & M\w{D}\wdiag}.
\end{align}
As three pairs of degenerate Weyl fermions form three Dirac neutrinos, the Dirac neutrino model is reproduced as expected.

@@

The matrix $U'$ is theoretically given with some matrix $\Theta$ by~\cite{Blennow:2011vn}\footnote{%
\TODO{}proof? or obvious if considers $\gSU(6)/(\gSU(3)\times\gSU(3))$?}
\begin{equation}
 U'\equiv \exp\pmat{0 & -\Theta^\dagger \\ \Theta & 0}
=\sum_{n=0}^\infty\pmat{
  {(-\Theta^\dagger\Theta)^{n}}/{(2n)!} &
  {-(-\Theta^\dagger\Theta)^{n}\Theta^\dagger}/{(2n+1)!}\\
  {+\Theta(-\Theta^\dagger\Theta)^{n}}/{(2n+1)!}&
  {(-\Theta\Theta^\dagger)^{n}}/{(2n)!}
}
\end{equation}
but practically we derive it in perturbative expansion.
For example, the type-I case is evaluated as\footnote{Cf.~\texttt{calculator/neutrino/diagonalization\_perturbative.wl}}
\begin{align}
 \text{(diag.)} = U^\TT\pmat{0 & M\w{D}^\TT \\ M\w{D} & M_N} U;\qquad
&U'\approx\pmat{
  1-\frac12 \Theta_0^\dagger \Theta_0&
  -\Theta_0^\dagger\\
 \Theta_0&
  1-\frac12\Theta_0\Theta_0^\dagger
};\quad \Theta_0:=-M_N^{-1}M\w{D},
\label{eq:typeIseesaw-blockdiag-1}\\
&\text{(block diag.)}=U'^\TT\pmat{0 & M\w{D}^\TT \\ M\w{D} & M_N} U'\approx\pmat{
  -M\w{D}^\TT M_N^{-1}M\w{D} & 0 \\ 0 & M_N
},
\label{eq:typeIseesaw-blockdiag-2}
\end{align}
where however note that $\Theta_0$ is close but not equal to $\Theta$.

@@

\subsection{PMNS matrix}\label{sec:PMNS}
As experiments have not found any deviation from $[U]\w{upper left}$ being unitary, we express the mixing with the PMNS matrix
 \GRAY{(Pontecorvo--\JAPANESE{牧}--\JAPANESE{中川}--\JAPANESE{坂田})}
parameterized by
\begin{equation}
\begin{split}
   U\w{PMNS}&=
 \pmat{1&&\\&\co{23}&\si{23}\\&-\si{23}&\co{23}}
 \pmat{\co{13}&&\si{13}\ee^{-\ii\delta\w{CP}}\\&1&\\-\si{13}\ee^{\ii\delta\w{CP}}&&\co{13}}
 \pmat{\co{12}&\si{12}&\\-\si{12}&\co{12}&\\&&1}
\pmat{\ee^{\ii\eta_1} \\ & \ee^{\ii\eta_2}\\&&1}
 %
\\&=
 \pmat{
 \co{12} \co{13} & \si{12} \co{13} & \si{13} \ee^{-\ii\delta\w{CP}}\\
 -\si{12} \co{23} - \co{12} \si{23} \si{13} \ee^{\ii\delta\w{CP}}& \co{12} \co{23} - \si{12}\si{23}\si{13} \ee^{\ii\delta\w{CP}}& \si{23}\co{13}\\
  \si{12}\si{23} - \co{12} \co{23} \si{13}\ee^{\ii\delta\w{CP}} & -\co{12}\si{23}-\si{12}\co{23}\si{13}\ee^{\ii\delta\w{CP}} & \co{23} \co{13}
}\pmat{\ee^{\ii\eta_1} \\ & \ee^{\ii\eta_2}\\&&1},
\end{split}
\end{equation}
where $\theta_{ij}\in[0,\pi/2]$ and $\delta\w{CP}\in[0,2\pi]$ as shown in \cref{eq:GeneralUnitary33}.
In the Dirac neutrino model, $\eta_1$ and $\eta_2$ (``Majorana phases'') are unphysical because we can rotate $\nR$ to eliminate them, while the rotation is not allowed in presence of the Majorana mass term.

@@

It also agrees with NuFIT~\cite[v5.0]{NUFIT} (their convention is given in Ref.~\cite{Esteban:2018azc}) except for the Majorana phases,
$
 \eta_i = \C{\alpha_i}.
$

\subsection{Casas--Ibarra parameterization}
We start from the approximation given in Eq.~\eqref{eq:typeIseesaw-blockdiag-2}:
\begin{equation}
 m\w L \approx -U\w{PMNS}^\TT M\w{D}^\TT M_N^{-1}M\w{D} U\w{PMNS},\qquad
 m\w H \approx U_X^\TT M_N U_X,
\end{equation}
where $m\w L$ and $m\w H$ are diagonal mass matrices for three light neutrinos and heavier neutrinos.
We then obtain
\begin{equation}
  m\w L
\approx -U\w{PMNS}^\TT M\w{D}^\TT (U_X^* m\w H U_X^\dagger)^{-1}M\w{D} U\w{PMNS}
=       -U\w{PMNS}^\TT M\w{D}^\TT U_Xm\w H^{-1}U_X^\TT M\w{D} U\w{PMNS}
\end{equation}
and decompose it as
\begin{equation}
   [\ii m\w L]^{1/2}[\ii m\w L]^{1/2}
\approx     \left(\frac{v}{\sqrt2}m\w H^{-1/2}U_X^\TT Y_n U\w{PMNS}\right)^{\TT} \left(\frac{v}{\sqrt2} m\w H^{-1/2}U_X^\TT Y_n U\w{PMNS}\right),
\end{equation}
which is the master equation for Casas-Ibarra parameterization~\cite{Casas:2001sr}.
Note that $U_X$ is unphysical, i.e., we can fix the basis of $\nR$ so that $U_X=1$.
This basis is equivalent (under our approximation) to the basis in which $M_N$ becomes diagonal.

\paragraph{Example: three right-handed neutrinos}
Let us assume all the neutrinos are massive thanks to three right-handed neutrinos. Then $M\w L\wdiag$ is invertible and
\begin{equation}
 R:=-\ii m\w H^{-1/2}U_X^\TT M\w{D}  U\w{PMNS}  m\w L^{-1/2}
\quad\Longrightarrow\quad R^\TT R\approx1.
\end{equation}
Conversely, with a matrix $R$ satisfying $R^\TT R=1$, the Yukawa matrix is given by
\begin{equation}
 \frac{v}{\sqrt2}Y_n
  \approx \ii U_X^*\sqrt{m\w H}R\sqrt{m\w L}U\w{PMNS}^\dagger
\end{equation}
or
\begin{equation}
 \frac{v}{\sqrt2}Y_n\Big|\w*{$M_N$ being diagonal}
  \approx \ii \sqrt{m\w H}R\sqrt{m\w L}U\w{PMNS}^\dagger.
\end{equation}

@@

With this parameterization, the neutrino mixing matrix is given by, in the basis with $M_N$ being diagonal,
\begin{equation}
 U \approx \pmat{
  U\w{PMNS}
  &  M\w{D}^\dagger M_N^{*-1} U_X
  \\ -M_N^{-1} M\w{D} U\w{PMNS}
  &  U_X
}
\approx\pmat{U\w{PMNS}&0\\0&1}\pmat{
  1
  &  -(-\ii{m\w H^{-1/2}}R{m\w L^{1/2}})^\dagger
  \\ -\ii m\w H^{-1/2}R{m\w L^{1/2}}
  &  1
}.
\end{equation}
